---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala, Rasika Nilatkar, Sayyam Palrecha, Hussain Nathani"
#date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```
#Outliers
```{r}
outlierKD2 <- function(df, var, rm=FALSE, boxplt=FALSE, histogram=TRUE, qqplt=FALSE) {
  dt = df # duplicate the dataframe for potential alteration
  var_name <- eval(substitute(var),eval(dt))
  na1 <- sum(is.na(var_name))
  m1 <- mean(var_name, na.rm = T)
  colTotal <- boxplt+histogram+qqplt
  par(mfrow=c(2, max(2,colTotal)), oma=c(0,0,3,0)) # fixed issue with only 0 or 1 chart selected
  if (qqplt) {
    qqnorm(var_name, main = "With outliers")
    qqline(var_name)
  }
  if (histogram) { hist(var_name, main="With outliers", xlab=NA, ylab=NA) }
  if (boxplt) { boxplot(var_name, main="With outliers") }

  outlier <- boxplot.stats(var_name)$out
  mo <- mean(outlier)
  var_name <- ifelse(var_name %in% outlier, NA, var_name)
  if (qqplt) {
    qqnorm(var_name, main = "Without outliers")
    qqline(var_name)
  }
  if (histogram) { hist(var_name, main="Without outliers", xlab=NA, ylab=NA) }
  if (boxplt) { boxplot(var_name, main="Without outliers") }
  
  if(colTotal > 0) {  # if no charts are wanted, skip this section
    title("Outlier Check", outer=TRUE)
    na2 <- sum(is.na(var_name))
    cat("Outliers identified:", na2 - na1, "\n")
    cat("Propotion (%) of outliers:", round((na2 - na1) / sum(!is.na(var_name))*100, 1), "\n")
    cat("Mean of the outliers:", round(mo, 2), "\n")
    m2 <- mean(var_name, na.rm = T)
    cat("Mean without removing outliers:", round(m1, 2), "\n")
    cat("Mean if we remove outliers:", round(m2, 2), "\n")
  }

  # response <- readline(prompt="Do you want to remove outliers and to replace with NA? [yes/no]: ")
  # if(response == "y" | response == "yes"){
  if(rm){
      dt[as.character(substitute(var))] <- invisible(var_name)
      #assign(as.character(as.list(match.call())$dt), dt, envir = .GlobalEnv)
      cat("Outliers successfully removed", "\n")
      return(invisible(dt))
  } else {
      cat("Nothing changed", "\n")
      return(invisible(df))
  }
}
```


# EDA

# Loading the Data
```{r}
flight <- data.frame(read.csv("Airline_Delay_Cause.csv"))
```

Dataset loaded..

Total number of rows: `r nrow(flight)`

Columns Description:
```{r}
d_list <- list(
  Column_name = c('year','month','carrier','carrier_name','airport','airport_name','arr_flights','arr_del15','carrier_ct','weather_ct','nas_ct','security_ct','late_aircraft_ct','arr_cancelled','arr_diverted','arr_delay','carrier_delay','weather_delay','nas_delay','security_delay','late_aircraft_delay'),
  description = c('Year data collected','Numeric representation of the month','Carrier.','Carrier Name.','Airport code.','Name of airport.','Number of flights arriving at airport','Number of flights more than 15 minutes late','Number of flights delayed due to air carrier. (e.g. no crew)','Number of flights due to weather.','Number of flights delayed due to National Aviation System (e.g. heavy air traffic).','Number of flights canceled due to a security breach.','Number of flights delayed as a result of another flight on the same aircraft delayed','Number of cancelled flights','Number of flights that were diverted','Total time (minutes) of delayed flight.','Total time (minutes) of delay due to air carrier','Total time (minutes) of delay due to inclement weather.','Total time (minutes) of delay due to National Aviation System.','Total time (minutes) of delay as a result of a security issue .','Total time (minutes) of delay flights as a result of a previous flight on the same airplane being late.') )
description <- as.data.frame(d_list)


#tab <- as.table(as.matrix(description))
#| tbl-colwidths: [20,80]
knitr::kable(description)
```

null rows in each column: `r colSums(is.na(flight))`

`r flight <- na.omit(flight)`

After nulls removal:

null rows after omiting nulls in each column: `r colSums(is.na(flight))`

Total number of rows: `r nrow(flight)`

Let's convert year, month, carrier and carrier names columns to factor variables.
``` {r}
library(data.table)

flight$year <- factor(flight$year)
flight$month <- factor(flight$month)
flight$carrier <- factor(flight$carrier)
flight$carrier_name <- factor(flight$carrier_name)

unique_carriers <- data.frame(unique(flight[,4]))
```


Summary of the Data: 
```{r}
knitr::kable(summary(flight))
```

List of carriers in the data:
```{r} 
knitr::kable(unique_carriers)
```




Scatter plots:
```{r}
library("dplyr")
flight_f <- flight[c('year','month','carrier','carrier_name','arr_flights','arr_delay','carrier_delay','weather_delay','nas_delay','security_delay','late_aircraft_delay')]
```



Outlier checks:

```{r}
par(mfrow = c(2, 3))
boxplot(flight_f$arr_delay,main = "Total Delay")
boxplot(flight_f$carrier_delay,main = "Delay as a result of air carrier")
boxplot(flight_f$weather_delay,main = "Delay as a result of inclement weather")
boxplot(flight_f$nas_delay,main = "Delay as a result of National Aviation System")
boxplot(flight_f$security_delay,main = "Delay as a result of security issue")
boxplot(flight_f$late_aircraft_delay,main = "Delay as a result of late previous flight")
```

Let's remove Outliers:

```{r}
flight_f <- outlierKD2(flight_f,arr_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
flight_f <- outlierKD2(flight_f,carrier_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
flight_f <- outlierKD2(flight_f,weather_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
flight_f <- outlierKD2(flight_f,nas_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
flight_f <- outlierKD2(flight_f,security_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
flight_f <- outlierKD2(flight_f,late_aircraft_delay,rm=FALSE, boxplt=FALSE, histogram=FALSE, qqplt=FALSE)
```

After outliers removal: 

```{r}
par(mfrow = c(2, 3))
boxplot(flight_f$arr_delay,main = "Total Delay")
boxplot(flight_f$carrier_delay,main = "Delay as a result of air carrier")
boxplot(flight_f$weather_delay,main = "Delay as a result of inclement weather")
boxplot(flight_f$nas_delay,main = "Delay as a result of National Aviation System")
boxplot(flight_f$security_delay,main = "Delay as a result of security issue")
boxplot(flight_f$late_aircraft_delay,main = "Delay as a result of late previous flight")
flight_s <- select_if(flight_f,is.numeric)
```

```{r}
pairs(flight_s, col='orange', main = "Pair plots for Flights data",cex.labels = 0.5,pch = 15,cex = 0.3)
```

```{r}
library(corrplot)
corr_matrix <- cor(flight_s)
corrplot(corr_matrix, method = "color",addCoef.col = "black", type = "lower", col = colorRampPalette(c("red", "yellow", "white", "blue","green"))(100),tl.col = "black", tl.srt= 5, main = "Correlation plot for Flights data",number.cex = 0.5)
```

Delay due to air carrier, delay flights as a result of a previous flight on the same airplane being late
