---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala, Rasika Nilatkar, Sayyam Palrecha, Hussain Nathani"
#date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

# SMART Question
**Which U.S. airports consistently experience the highest cumulative delays across different delay causes (Carrier, Weather, NAS, Security, Late Aircraft) for both origin and destination flights, and how does this ranking vary on a monthly or seasonal basis over the observed period?**

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(echo = T, warning = F, results = "markup", message = F)
options(scientific=T, digits = 3)  
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

### Understanding the dataset
```{r Initial dataset, include=T}
flight <- data.frame(read.csv("Flight_delay.csv"))
str(flight)
```

### Necessary Libraries
```{r Libraries, include=T}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
```

### Filtering Data
```{r Filtering data, include=T}
flight_data <- flight %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay))
```

### Formatting Date and adding Month and Season column
```{r Formatting data, include=T}
flight_data$Date <- as.Date(flight_data$Date, format = "%d-%m-%Y")
flight_data_cleaned <- flight_data %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE),
         Season = case_when(
           Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
           Month %in% c("Mar", "Apr", "May") ~ "Spring",
           Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
           Month %in% c("Sep", "Oct", "Nov") ~ "Fall"
         ))
```

### Cleaning Dataframe
```{r Cleaning data, include=T}
flight_delays <- flight_data_cleaned %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay)) %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Month, Season)
```

### Summary of cleaned data
```{r Summary, include=T}
knitr::kable(summary(flight_delays))
```

### Checking NA count
```{r Checking NA values, include=T} 
na_count <- colSums(is.na(flight_delays))
print(na_count)
```

### Head of Cleaned data
```{r Cleaned data, include=T}
head(flight_delays)
```

### Checking all the rows if delay exists or not
```{r Delay checks, include=T}
flight_delays$outcome <- ifelse(rowSums(flight_delays == 0) == ncol(flight_delays), 1, 0)
no_delay_rows <- flight_delays[flight_delays$outcome == 1, ]
head(no_delay_rows)
```

### Frequency of top 10 Origin and Destination airports delay
```{r Identifying frequency, include=T}
origin_counts <- flight_delays %>%
  group_by(Origin) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  slice(1:10) %>%
  mutate(Type = "Origin")

dest_counts <- flight_delays %>%
  group_by(Dest) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  slice(1:10) %>%
  mutate(Type = "Destination")

origin_counts <- rename(origin_counts, Airport = Origin)
dest_counts <- rename(dest_counts, Airport = Dest)
combined_counts <- bind_rows(origin_counts, dest_counts)

ggplot(combined_counts, aes(x = reorder(Airport, Frequency), y = Frequency, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 10 Airports by Delay Frequency (Origin vs Destination)",
       x = "Airport", y = "Delay Frequency") +
  theme_minimal() +
  coord_flip() +
  scale_fill_manual(values = c("Origin" = "steelblue", "Destination" = "darkorange"))
```

### Creating Heatmap to understand delay cause at top 10 airports
```{r Identifying delay causes, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

heatmap_data <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay) %>%
  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType",
               values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  filter(!is.na(TotalDelay))

heatmap_summary <- heatmap_data %>%
  group_by(Airport, DelayType) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE)) %>%
  ungroup()

ggplot(heatmap_summary, aes(x = Airport, y = DelayType, fill = TotalDelay)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "white", high = "blue", na.value = "grey50") +
  labs(title = "Heatmap of Delay Causes at Top 10 Airports",
       x = "Airport",
       y = "Delay Cause",
       fill = "Total Delay Minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Identifying season in which delays occurs the most at top 10 airports
```{r Identifying seasons, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

print(colnames(filtered_delays))

seasonal_delays <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Season) %>%

  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType", values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  group_by(Airport, Season) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE), .groups = 'drop')

ggplot(seasonal_delays, aes(x = Airport, y = TotalDelay, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Seasonal Delays for Top 10 Airports",
       x = "Airport", y = "Total Delay (Minutes)") +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set3")
```

# Result Summary

The analysis reveals that among U.S. airports, **O'Hare International (ORD), Hartsfield-Jackson Atlanta International (ATL), and Dallas/Fort Worth International (DFW)** consistently rank as the top three airports with the highest cumulative delays across various causes. The primary contributors to these delays are Late Aircraft Delays, NAS Delays, and Carrier Delays, whereas Weather Delays occur less frequently, and Security Delays are rare. Additionally, delays peak during the Winter season, with Spring and Summer following in frequency, indicating a seasonal pattern that could aid in better resource allocation and management at these airports.