---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala, Rasika Nilatkar, Sayyam Palrecha, Hussain Nathani"
#date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

# SMART Question
**What are the significant factors influencing total flight delays across the top 10 busiest U.S. airports over the observed period, and how do these factors vary by season?**

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(echo = T, warning = F, results = "markup", message = F)
options(scientific=T, digits = 3)  
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

### Understanding the dataset
```{r Initial dataset, include=T}
flight <- data.frame(read.csv("Flight_delay.csv"))
str(flight)
```

### Necessary Libraries
```{r Libraries, include=T}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
```

### Filtering Data
```{r Filtering data, include=T}
flight_data <- flight %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay))
```

### Formatting Date and adding Month and Season column
```{r Formatting data, include=T}
flight_data$Date <- as.Date(flight_data$Date, format = "%d-%m-%Y")
flight_data_cleaned <- flight_data %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE),
         Season = case_when(
           Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
           Month %in% c("Mar", "Apr", "May") ~ "Spring",
           Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
           Month %in% c("Sep", "Oct", "Nov") ~ "Fall"
         ))
```

### Cleaning Dataframe
```{r Cleaning data, include=T}
flight_delays <- flight_data_cleaned %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay)) %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Month, Season)
```

### Summary of cleaned data
```{r Summary, include=T}
knitr::kable(summary(flight_delays))
```

### Checking NA count
```{r Checking NA values, include=T} 
na_count <- colSums(is.na(flight_delays))
print(na_count)
```

### Head of Cleaned data
```{r Cleaned data, include=T}
head(flight_delays)
```

### Checking all the rows if delay exists or not
```{r Delay checks, include=T}
flight_delays$outcome <- ifelse(rowSums(flight_delays == 0) == ncol(flight_delays), 1, 0)
no_delay_rows <- flight_delays[flight_delays$outcome == 1, ]
head(no_delay_rows)
```

### Frequency of top 10 Origin and Destination airports delay
```{r Identifying frequency, include=T}
origin_counts <- flight_delays %>%
  group_by(Origin) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  mutate(Type = "Origin")

dest_counts <- flight_delays %>%
  group_by(Dest) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  mutate(Type = "Destination")

origin_counts <- rename(origin_counts, Airport = Origin)
dest_counts <- rename(dest_counts, Airport = Dest)

combined_counts <- bind_rows(origin_counts, dest_counts)

top_airports <- combined_counts %>%
  group_by(Airport) %>%
  summarise(TotalFrequency = sum(Frequency)) %>%
  arrange(desc(TotalFrequency)) %>%
  slice(1:10)

final_counts <- combined_counts %>%
  filter(Airport %in% top_airports$Airport)

ggplot(final_counts, aes(x = reorder(Airport, Frequency), y = Frequency, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 10 Airports by Delay Frequency (Origin vs Destination)",
       x = "Airport", y = "Delay Frequency") +
  theme_minimal() +
  coord_flip() +
  scale_fill_manual(values = c("Origin" = "steelblue", "Destination" = "darkorange"))
```

### Creating Heatmap to understand delay cause at top 10 airports
```{r Identifying delay causes, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

heatmap_data <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay) %>%
  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType",
               values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  filter(!is.na(TotalDelay))

heatmap_summary <- heatmap_data %>%
  group_by(Airport, DelayType) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE)) %>%
  ungroup()

ggplot(heatmap_summary, aes(x = Airport, y = DelayType, fill = TotalDelay)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "white", high = "blue", na.value = "grey50") +
  labs(title = "Heatmap of Delay Causes at Top 10 Airports",
       x = "Airport",
       y = "Delay Cause",
       fill = "Total Delay Minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Identifying season in which delays occurs the most at top 10 airports
```{r Identifying seasons, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

print(colnames(filtered_delays))

seasonal_delays <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Season) %>%

  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType", values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  group_by(Airport, Season) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE), .groups = 'drop')

ggplot(seasonal_delays, aes(x = Airport, y = TotalDelay, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Seasonal Delays for Top 10 Airports",
       x = "Airport", y = "Total Delay (Minutes)") +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set3")
```

### Two-Sample T-Test: Comparing Delays Between Two Airports
```{r T-test, include=T}
airport_data <- flight_delays %>% filter(Origin %in% c("ORD", "DFW"))

t_test_result <- t.test(
  LateAircraftDelay ~ Origin,
  data = airport_data,
  alternative = "two.sided")

print(t_test_result)
```

### ANOVA: Comparing Mean Delays Across Seasons
```{r Anova-test, include=T}
anova_result <- aov(CarrierDelay ~ Season, data = flight_delays)
summary(anova_result)
```

### Tukey HSD test for identifying which pairs of seasons differ significantly
```{r Tukey-test, include=T}
tukey_result <- TukeyHSD(anova_result)

print(tukey_result)
```

### Linear Regression Model to Analyze Total Flight Delays Across Airports and Seasons
```{r LM, include=T}
flight_data_cleaned <- seasonal_delays %>%
  group_by(Airport, Season) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE), .groups = 'drop')

flight_data_cleaned$Season <- factor(flight_data_cleaned$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

model <- lm(TotalDelay ~ Airport + Season, data = flight_data_cleaned)

xkabledply(model, title = paste("Model :", format(formula(model)) ) )
summary(model)
```

```{r}
predictions <- predict(model, newdata = flight_data_cleaned)

predictions
```

### Visualizing Predictions versus Actual values
```{r Predictions, include=T}
plot(flight_data_cleaned$TotalDelay, type = "p", col = "blue", pch = 16, cex = 1.2,
     xlab = "Observation Index", ylab = "TotalDelay",
     main = "Comparison of Actual and Predicted Total Delays")

points(predictions, col = "red", pch = 17, cex = 1.2)

legend("topright", legend = c("Actual", "Predicted"),
       col = c("blue", "red"), pch = c(16, 17), cex = 0.8)
```


```{r}
par(mfrow = c(2, 2))
plot(model)
```

```{r}
mse <- mean(model$residuals^2)
print(mse)

rmse <- sqrt(mse)
print(rmse)
```

# Final Conclusion:

1. Key Factors Influencing Flight Delays:
Late Aircraft Delays, NAS Delays, and Carrier Delays are the primary contributors to total flight delays. Weather-related delays occur less frequently, while Security Delays are rare.
Airports such as O'Hare International (ORD), Dallas/Fort Worth International (DFW), and Hartsfield-Jackson Atlanta International (ATL) consistently experience the highest cumulative delays due to these factors.

2. Seasonal Variations:
The Winter season experiences higher delays compared to other seasons, with Summer seeing fewer delays, as confirmed by the ANOVA test and the Tukey HSD test.

3. Airport-Specific Findings:
O'Hare International (ORD) and Dallas/Fort Worth International (DFW) are outliers with consistently higher delays, as shown by the T-test results.
On the other hand, airports like JFK and LGA experience lower delays, making them more efficient in terms of total delays compared to others.

4. Linear Regression Model Insights:
Airports like JFK and LGA experience significantly lower delays compared to the baseline, while ORD experiences higher delays.
Season is a significant predictor of delays, with Summer having fewer delays than Winter. The model's Multiple R-squared value of 0.891 indicates a strong fit, suggesting that these factors together explain a significant portion of the variation in total delays.

5. Model Performance:
The Mean Squared Error (MSE) of 3.58e+10 and Root Mean Squared Error (RMSE) of 189,221 indicate a reasonable fit, with the model performing adequately in predicting total flight delays across airports and seasons.