---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala,  Hussain Nathani, Rasika Nilatkar, Sayyam Palrecha"
#date: "today"
date: "`r Sys.Date()`"
subtitle: "Team 2 - JamesRiverhouse"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r Libraries, include=T}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(corrplot)
library(scales)
library(gridExtra)
library(stats)
library(reshape2)
```

# Introduction:

According to a market report by AirHelp, in 2022, at least 200 million passengers were affected by delays and cancellations in the US, which impacted on the economy of $30–34 billion in the US. With the increasing volume of air traffic, especially during peak travel seasons, understanding the causes and patterns of delays is crucial for airlines and airports to optimize resources and improve performance. This project aims to predict and analyze U.S. flight delays using data-driven techniques in R, focusing on different causes of delays, seasonal variations, and time-of-day effects. By answering key questions, we can provide insights that help airlines and airports mitigate delays and enhance operational efficiency.

Previous research on flight delays has focused on identifying key causes and attempting to forecast delays based on historical data. Studies often show weather as a major contributor, particularly in seasonal peaks, and highlight airports with chronic delays. Many existing analyses have also sought to classify delays by type, allowing airlines to target specific operational changes.

Exploring the dataset collected from Kaggle allows us to see delay patterns directly from the data. By analyzing patterns in delay causes and seasonality, we can make actionable recommendations. Our findings provide airlines and airport management with insights that could lead to better resource allocation and proactive delay management, ultimately improving the traveler experience and operational efficiency.



# 1. Loading the Dataset
```{r}
flight <- data.frame(read.csv("Flight_delay.csv"))
str(flight)
```

Dataset loaded..

```{r}
column_names <- c('Day_of_Week','Date','Departure_Time','Arrival_Time','Scheduled_Arrival_Time',  'Unique_Carrier_Code','Carrier_name', 'Flight_Number','Tail_Number', 'Actual_Elapsed_Time_min','Estimated_Elapsed_Time_min','Air_Time_min', 'Arrival_Delay_min', 'Departure_Delay_min','Origin', 'Origin_Airport','Destination','Destination_Airport', 'Distance_mi','Taxi_In_min','Taxi_Out_min', 'Cancelled', 'CancellationCode','Diverted','Carrier_Delay_min','Weather_Delay_min', 'NAS_Delay_min','Security_Delay_min','Late_Aircraft_Delay_min')

colnames(flight) <- column_names
unique_carriers <- data.frame(unique(flight[,7]))
```


# 2. Data Cleaning & Summary of Dataset

The dataset from the Kaggle link titled "Flight Delay and Causes" contains
comprehensive information on U.S. flight delays, focusing on various causes of delays
and their impacts across different airlines and airports. The data is sourced from the
U.S. Department of Transportation's Bureau of Transportation Statistics (BTS) and
includes detailed records of flight delays, cancellations, and their contributing factors.

Summary:
```{r}
summary(flight)
```

Number of rows before Data pre-processing: `r nrow(flight)`

The airlines considered for analysis are:

`r knitr::kable(unique_carriers)`


Reasons for the delay:

1. CarrierDelay → Flight delay due to carrier(e.g. maintenance or crew problems, aircraft cleaning, fueling, etc), 0 = No, yes = (in minutes)

2. WeatherDelay → Flight delay due to weather, 0 = No, yes = (in minutes)

3. NASDelay → Flight delay by NSA(National Aviation System), 0 = No, yes = (in minutes)

4. SecurityDelay → Flight delay by security reason, 0 = No, yes = (in minutes)

5. LateAircraftDelay → Flight delay by Aircraft reason, 0 = No, yes = (in minutes)


# 3. Data Preprocessing

## 3.1. Outliers and null value removal

Outlier removal improves data analysis accuracy by eliminating extreme values that can skew statistical measures and models. This leads to more reliable and interpretable results.

```{r}
remove_outliers <- function(df, columns) {
    df_clean <- df
    
    for(column_name in columns) {
        # Get the specified column
        x <- df_clean[[column_name]]
        
        # Removing NA values
        valid_rows <- !is.na(x)
        x_clean <- x[valid_rows]
        
        # Calculating Q1, Q3, and IQR
        Q1 <- quantile(x_clean, 0.25)
        Q3 <- quantile(x_clean, 0.75)
        IQR <- Q3 - Q1
        
        # bounds for outlier removal
        lower_bound <- Q1 - 1.5 * IQR
        upper_bound <- Q3 + 1.5 * IQR
        
        # summary before removal
        cat("\nProcessing", column_name, ":\n")
        cat("Original rows:", sum(valid_rows), "\n")
        cat("Q1:", Q1, "Q3:", Q3, "IQR:", IQR, "\n")
        cat("Lower bound:", lower_bound, "Upper bound:", upper_bound, "\n")
        
        # Deleting outlier rows
        keep_rows <- x >= lower_bound & x <= upper_bound & valid_rows
        df_clean <- df_clean[keep_rows, ]
        
        # summary after removal
        cat("Rows after removal:", nrow(df_clean), "\n")
        cat("Outliers removed:", sum(valid_rows) - nrow(df_clean), "\n\n")
    }
    
    return(df_clean)
}
```



```{r}
library(ggplot2)
flight_b <- remove_outliers(flight,c('Arrival_Delay_min','Carrier_Delay_min','Weather_Delay_min','NAS_Delay_min','Security_Delay_min','Late_Aircraft_Delay_min'))
```

Number of rows after nulls and Outlier removal: `r nrow(flight_b)`

## 3.2. Normality checks

Parametrical statistical testing can be done if the data points are normal. Let's check.


```{r}
#| fig-width: 20
#| fig-height: 15
par(mfrow = c(3, 2))
qqnorm(flight_b$Carrier_Delay_min, ylab = 'Flight delay due to carrier', main = "Flight delay due to carrier"); qqline(flight_b$Carrier_Delay_min,col=2)
qqnorm(flight_b$Weather_Delay_min, ylab = 'Flight delay due to weather', main = "Flight delay due to weather"); qqline(flight_b$Weather_Delay_min,col=2)
qqnorm(flight_b$NAS_Delay_min, ylab = 'NAS_Delay_min', main = "Flight delay by NSA(National Aviation System)"); qqline(flight_b$NAS_Delay_min,col=2)
qqnorm(flight_b$Security_Delay_min, ylab = 'Security_Delay_min', main = "Flight delay by security reason"); qqline(flight_b$Security_Delay_min,col=2)
qqnorm(flight_b$Late_Aircraft_Delay_min, ylab = 'Flight delay by Aircraft reason', main = "Flight delay by Aircraft reason"); qqline(flight_b$Late_Aircraft_Delay_min,col=2)
qqnorm(flight_b$Arrival_Delay_min, ylab = 'Total Delay' , main = "Total Delay"); qqline(flight_b$Arrival_Delay_min,col=2)

```


weather and security delays are rare in this dataset and when they occur, they're likely significant outliers.

None of the delay types follow a normal distribution.

Since these distributions are not normal, Parametric statistical studies cannot be performed. 


# 4. SMART Questions

## SMART Q1
"Can we predict the total arrival delay using ML modeling for the top 5 most delayed airlines, using key factors that holds impact on the delay. This helps identify the significant predictors of delays and provides actionable insights for resource allocation and operational planning in future from the data collected from January 2019 to June 2019, catering better customer satisfaction."

### 4.1. Identifying top 5 airlines:

To Identify top 5 airlines, comparing median and mean values on boxplots would help.


```{r}
#| fig-width: 20
#| fig-height: 5
ggplot(flight_b,aes(Carrier_name,Arrival_Delay_min))+geom_boxplot(aes(fill = Carrier_name ))+labs(title = "Box plots of Airlines flight delay distribution") + theme(plot.title = element_text(size = 25,hjust =0.5)) + stat_summary(aes(y = Arrival_Delay_min , label = round(..y.., 2)), 
                 fun.y = median, geom = "text") 
```



From the box-plots above, higher median lines are found for:

JetBlue Airways

United Air Lines Inc.

American Eagle Airlines Inc.

American Airlines Inc.

US Airways Inc.


```{r}
flight_5 <- flight_b[flight_b$Carrier_name %in% c('JetBlue Airways','United Air Lines Inc.','US Airways Inc.','American Airlines Inc.','American Eagle Airlines Inc.'),]
```

```{r}
library(dplyr)

# Adding the number of flights on a date level for each origin
flight_b <- flight_b %>%
  group_by(Date, Origin) %>%
  mutate(Flights_At_Origin = n()) %>%
  ungroup()

# Adding the number of flights on a date level for each destination
flight_b <- flight_b %>%
  group_by(Date, Destination) %>%
  mutate(Flights_At_Destination = n()) %>%
  ungroup()

```

### 4.2. Identifying major reasons for these 5 airlines:

If we can understand the major reasons impacting top 5 delayed airlines, that would cover majority of the delayed reasons in the US according to the available dataset.


```{r}
library(ggplot2)
library(reshape2)
avg_delays <- aggregate(
  cbind(
    Carrier_Delay_min, 
    Weather_Delay_min, 
    NAS_Delay_min, 
    Security_Delay_min, 
    Late_Aircraft_Delay_min
  ) ~ Carrier_name, 
  data = flight_5, 
  FUN = mean, 
  na.rm = TRUE
)

delay_long <- melt(
  avg_delays,
  id.vars = "Carrier_name"
)

delay_long <- melt(avg_delays, id.vars = "Carrier_name")
delay_totals <- aggregate(value ~ Carrier_name, data = delay_long, sum)
delay_long <- merge(delay_long, delay_totals, by = "Carrier_name", 
                   suffixes = c("", "_total"))
delay_long$percentage <- (delay_long$value / delay_long$value_total)*100

threshold <- 0
delay_long$label <- ifelse(delay_long$percentage >  threshold, sprintf("%.1f%%",delay_long$percentage),"") 

```


```{r}
ggplot(delay_long, aes(x = Carrier_name, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = label),
            position = position_fill(vjust = 0.5),size = 3)+
  theme_minimal() +
  labs(
    title = "Proportion of Delay Types by Airline",
    x = "Airline",
    y = "Percentage",
    fill = "Delay Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +  
  scale_y_continuous(labels = scales::percent)

flight_5$MonthYear <- format(as.Date(flight_5$Date, "%d-%m-%Y"), "%b-%Y")

# Calculate monthly averages for each airline and delay type
monthly_delays <- aggregate(
  cbind(
    Carrier_Delay_min, 
    Weather_Delay_min, 
    NAS_Delay_min, 
    Security_Delay_min, 
    Late_Aircraft_Delay_min
  ) ~ MonthYear + Carrier_name, 
  data = flight_5, 
  FUN = mean, 
 na.rm = TRUE
)
```



```{r}
delay_long <- melt(monthly_delays, 
                  id.vars = c("MonthYear", "Carrier_name"),
                  variable.name = "Delay_Type",
                  value.name="Minutes")
```


Analysis from above graph:

1. Late Aircraft Delay dominates across all airlines, accounting for the largest proportion of delays, highest being in JetBlue Airways (64.0%).

2. Followed by Carrier Delay and NAS Delay, all three combined causes 100% delays.

3. Weather and Security delays appear minimal or non-existent

4. Late aircraft delays being dominant suggests a huge impact in airline operations

5. Different airlines might need different strategies based on their delay profiles, for example JetBlue might focus on late aircraft issues, while American Airlines might prioritize carrier-related delays since American Airlines has the highest percentage compared to others.


### 4.3. Identifying major reasons for these 5 airlines

To validate if the top5 reasons are major reasons impacting most of the flight delays, let's check the consistency of the reasons percentage across the available timeline.


```{r}
library(dplyr)

delay_long <- delay_long %>%
  mutate(
    MonthDate = as.Date(paste0("01-", MonthYear), format = "%d-%b-%Y"),
    MonthYear = factor(format(MonthDate, "%b-%Y"),levels = format(sort(unique(MonthDate)),"%b-%Y")))
ggplot(delay_long, 
       aes(x = MonthYear, 
           y = Minutes, 
           color = Delay_Type, 
           group = Delay_Type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Carrier_name, ncol = 2, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Monthly Delay Trends by Airline and Delay Type",
    x = "Month-Year",
    y = "Average Delay Minutes",
    color = "Delay Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    panel.spacing = unit(1, "lines")
  ) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(nrow=1))

```

Consistency of analysis over 6 month period:

1. Late Aircraft Delays (orange line) consistently show the highest delays across all airlines

2. Security Delays (purple line) consistently show the lowest delays

3. Most trends remain relatively stable across the 6-month period

### 4.4. Modelling:


#### 4.4.1. Data preparation for modelling.


```{r}
ds <- subset(flight_5,select = -c(Carrier_Delay_min,Origin_Airport,Destination_Airport,CancellationCode,MonthYear,Carrier_name,Weather_Delay_min,NAS_Delay_min,Security_Delay_min,Late_Aircraft_Delay_min))

ds$Cancelled <- as.factor(ds$Cancelled)
ds$Diverted <- as.factor(ds$Diverted)
ds$Flight_Number <- as.factor(ds$Flight_Number)
ds$Tail_Number <- as.factor(ds$Tail_Number)
```

There are no cancelled and diverted flights, hence let's drop the columns

```{r}
ds <- subset(ds,select = -c(Cancelled,Diverted))
```


Let's convert columns into factor variables

```{r}
ds$Day_of_Week <- as.factor(ds$Day_of_Week)
ds$Date <- as.Date(ds$Date, format = "%d-%m-%Y")

ds$Unique_Carrier_Code <- as.factor(ds$Unique_Carrier_Code)
ds$Origin <- as.factor(ds$Origin)
ds$Destination <- as.factor(ds$Destination)
```

`Arrival_Time`, `Scheduled_Arrilval_Time`, `Departure Time` are left in integer format, since the time of the day are ordinal variables and has clear order.


```{r}
ds <- subset(ds,select = -c(Flight_Number,Tail_Number))
```

#### 4.4.2. Features Exploration

Checking the features using forward selection method.

```{r}
loadPkg("leaps")
reg.forward10 <- regsubsets(Arrival_Delay_min~., data = ds, nvmax = 10, nbest = 2, method = "forward")
plot(reg.forward10, scale = "adjr2", main = "Adjusted R^2")
```
```{r}
reg_summary <- summary(reg.forward10)
best_model_index <- which.max(reg_summary$adjr2)  # Index of highest Adjusted R²
best_model_features <- reg_summary$which[best_model_index, ]  # Logical vector of selected features
selected_features <- names(best_model_features[best_model_features == TRUE])[-1]
cat("Features with the best Adjusted R²:\n")
cat(selected_features)
ds_filtered <- ds[, c("Arrival_Delay_min", selected_features)]
```
Since the features with best Adjusted $R^2$ are `Estimated_Elapsed_Time_min`, `Air_Time_min`, `Departure_Delay_min` , `Taxi_In_min` and `Taxi_Out_min`. Let's re-check Adjusted $R^2$ graph with just these metrics.

```{r}
loadPkg("leaps")
reg.forward10 <- regsubsets(Arrival_Delay_min~., data = ds_filtered, nvmax = 10, nbest = 2, method = "forward")
plot(reg.forward10, scale = "adjr2", main = "Adjusted R^2")
```

Let's build a model with these features.

```{r}
model <- lm(Arrival_Delay_min ~ Estimated_Elapsed_Time_min+Air_Time_min+Departure_Delay_min+Taxi_In_min+Taxi_Out_min, data = ds)
summary(model)
```

$R^2$ value of 1 may seem ideal, it’s rare in real-world datasets. This perfect fit could indicate issues such as:
Overfitting: The model may be too complex and tightly fit to the specific dataset.
Multicollinearity: There may be high correlations among the predictors, leading to redundancy.

Let's check multi collinearity for the model.
```{r}
xkablevif(model)
```
As expected, high multi - collinearity.

Hence, Let's build multiple models using the above variables.

First, let's check correlation of these variables with `Arrival_Delay_min`

```{r, fig.height=6, fig.width=15}
correlation_matrix <- cor(select_if(ds_filtered,is.numeric))
par(mar = c(0, 0, 0, 0), cex.main = 0.8, cex.axis = 0.8, cex.lab = 3.0)

corrplot(correlation_matrix, 
         method = "color", 
         addCoef.col = "black", 
         col = colorRampPalette(c("maroon", "yellow", "white", "lightblue", "palegreen"))(100), 
         tl.col = "black", 
         tl.cex = 0.8,          # Scale for text labels
         tl.srt = 30,           # Rotate labels for better spacing
         number.cex = 0.8,      # Scale for correlation coefficients
         mar = c(0, 0, 1, 0),   # Adjust margins (bottom, left, top, right)
         main = "Correlation plot for top 5 most delayed flight data")

```

`Departure_Delay_min` has high correlation with `Arrival_Delay_min`, which also makes sense because delay in departure also leads to flight reaching the destination late.

#### 4.4.3. Modelling
Model 1: 

```{r}
model1 <- lm(Arrival_Delay_min~Departure_Delay_min,data = ds_filtered)
summary(model1)
```
$R^2$ being 74.5% and the p-value suggests that `Departure_Delay_min` is a significant factor. The F-statistic is extremely large further confirming the model's overall significance.

The second highest correlated factor is `Taxi_Out_min`.

Model 2:

```{r}
model2 <- lm(Arrival_Delay_min~(Departure_Delay_min*Taxi_Out_min)^2,data = ds_filtered)
summary(model2)
```
$R^2$ being 80.3% with interaction terms and the p-value suggests that `Departure_Delay_min` and `Taxi_Out_min` are a significant factor. The F-statistic is extremely large further confirming the model's overall significance.

Let's check the VIF of model2

```{r}
xkablevif(model2)
```

Vif = 1.03, Hence there's no multicollinearity.


This suggests that the time elapsed between departure from the origin airport gate and wheels off is also a significant factor to the predict the flight delay.
Extended Taxi-Out time often occurs due to congestion at the runway or delays in the sequencing of departures.
Adverse weather (rainy/winters), Insufficient ground resources (as explored in the EDA).
Taxi-Out delays peaks during holiday seasons/ winter months, de-icing procedures can extend the Taxi-Out Time, contributing to departure and arrival delays.

Model 3:

```{r}
model3 <- lm(Arrival_Delay_min~(Departure_Delay_min+Taxi_Out_min+Taxi_In_min)^2,data = ds_filtered)
summary(model3)
```
There's a significant improvement in $R^2$ value (82.3%) and p-values suggest that the predictors are significant.
Same as Taxi_Out delays, Taxi_In delays also peaks during holiday seasons and winter/rainy seasons.

Let's check vif values.

```{r}
xkablevif(model3)
```

VIFs until 10 can be accepted.


Model 4:

```{r}
model4 <- lm(Arrival_Delay_min ~ (Departure_Delay_min + Taxi_Out_min + 
    Taxi_In_min+Air_Time_min)^2, data = ds_filtered)
summary(model4)
```
Not a significant impact on the $R^2$ values.
Let's check VIF values.

```{r}
xkablevif(model4)
```

Multi collinearity kicks in when`Air_Time_min` is added.

Model 5:

```{r}
model5 <- lm(Arrival_Delay_min ~ (Departure_Delay_min + Taxi_Out_min + 
    Taxi_In_min+Estimated_Elapsed_Time_min)^2, data = ds_filtered)
summary(model5)
```
$R^2$ didn't improve significantly, but let's check vif
```{r}
xkablevif(model5)
```

VIF crossing 10. Hence, multi collinearity is present.

#### 4.4.3. Model Assesment:

Now, let's compare model1, model2, model3 using ANOVA tests.

```{r}
anova_model <- anova(model1,model2,model3)
anova_model
```
Model Comparisons
Model 1 vs. Model 2:

Adding (Departure_Delay_min * Taxi_Out_min)^2 explains an additional 4,152,475 units of variance (Sum of Sq).
The F-statistic is 24,285, with a p-value < 2e-16, indicating this improvement is highly significant.
Model 2 vs. Model 3:

Adding Taxi_In_min and its interactions explains an additional 1,451,645 units of variance (Sum of Sq).
The F-statistic is 5,660, with a p-value < 2e-16, indicating this improvement is also highly significant.

Each step in model complexity significantly improves the fit, as shown by the large F-statistics and extremely small p-values (< 0.001).]The additional variance explained decreases from Model 2 to Model 3 (4,152,475 vs. 1,451,645). This suggests that while adding more terms improves the model, the magnitude of improvement is smaller.


Let's check MSE and RMSE for the model built using the features from the best model above:

```{r}
library(caTools)
set.seed(123) #for reproducability
split <- sample.split(ds_filtered$Arrival_Delay_min, SplitRatio = 0.8)
train_data <- subset(ds_filtered, split == TRUE)
test_data <- subset(ds_filtered, split == FALSE)
model2_train <- lm(Arrival_Delay_min ~ (Departure_Delay_min*Taxi_Out_min)^2, data = train_data)
model3_train <- lm(Arrival_Delay_min~(Departure_Delay_min+Taxi_Out_min+Taxi_In_min)^2, data = train_data)

predictions_2_test <- predict(model2_train, newdata = test_data)
predictions_3_test <- predict(model3_train, newdata = test_data)

predictions_2_train <- predict(model2_train, newdata = train_data)
predictions_2_train <- predict(model2_train, newdata = train_data)
```

```{r}
# Function to calculate MSE
calculate_mse <- function(actuals, predictions) {
  mean((actuals - predictions)^2)
}

# Function to calculate RMSE
calculate_rmse <- function(actuals, predictions) {
  sqrt(calculate_mse(actuals, predictions))
}

# Calculate MSE and RMSE for model2 (Train and Test)
mse_model2_train <- calculate_mse(train_data$Arrival_Delay_min, predictions_2_train)
rmse_model2_train <- calculate_rmse(train_data$Arrival_Delay_min, predictions_2_train)

mse_model2_test <- calculate_mse(test_data$Arrival_Delay_min, predictions_2_test)
rmse_model2_test <- calculate_rmse(test_data$Arrival_Delay_min, predictions_2_test)

# Calculate MSE and RMSE for model3 (Train and Test)
predictions_3_train <- predict(model3_train, newdata = train_data)  # Predictions for training data

mse_model3_train <- calculate_mse(train_data$Arrival_Delay_min, predictions_3_train)
rmse_model3_train <- calculate_rmse(train_data$Arrival_Delay_min, predictions_3_train)

mse_model3_test <- calculate_mse(test_data$Arrival_Delay_min, predictions_3_test)
rmse_model3_test <- calculate_rmse(test_data$Arrival_Delay_min, predictions_3_test)

# Print results
cat("Model 2 - Training MSE:", mse_model2_train, "RMSE:", rmse_model2_train, "\n")
cat("Model 2 - Testing MSE:", mse_model2_test, "RMSE:", rmse_model2_test, "\n\n")

cat("Model 3 - Training MSE:", mse_model3_train, "RMSE:", rmse_model3_train, "\n")
cat("Model 3 - Testing MSE:", mse_model3_test, "RMSE:", rmse_model3_test, "\n")

```
Both Model 2 and Model 3 demonstrate low MSE and RMSE values for both training and testing datasets, indicating they are well-fitted to the data and capable of making accurate predictions.
The RMSE values are close to each other for the training and testing sets, showing no signs of overfitting or underfitting.

The lower MSE and RMSE for Model 3 suggest it should be preferred for predicting Arrival_Delay_min.



# 5. SMART Questions

## SMART Q2
"What are the significant factors influencing total flight delays across the top 10 busiest U.S. airports over the observed period, and how do these factors vary by season?"

### 5.1 Question Specific Data Preprocessing
```{r Filtering data, include=T}
flight <- data.frame(read.csv("Flight_delay.csv"))

flight_data <- flight %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay))
```

#### 5.1.1 Formatting Date and adding Month and Season column
```{r Formatting data, include=T}
flight_data$Date <- as.Date(flight_data$Date, format = "%d-%m-%Y")
flight_data_cleaned <- flight_data %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE),
         Season = case_when(
           Month %in% c("Dec", "Jan", "Feb") ~ "Winter",
           Month %in% c("Mar", "Apr", "May") ~ "Spring",
           Month %in% c("Jun", "Jul", "Aug") ~ "Summer",
           Month %in% c("Sep", "Oct", "Nov") ~ "Fall"
         ))
```

#### 5.1.2 Cleaning Dataframe
```{r Cleaning data, include=T}
flight_delays <- flight_data_cleaned %>%
  filter(!is.na(CarrierDelay) | !is.na(WeatherDelay) | 
         !is.na(NASDelay) | !is.na(SecurityDelay) | 
         !is.na(LateAircraftDelay)) %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Month, Season)
```

#### 5.1.3 Summary of cleaned data
```{r Summary, include=T}
knitr::kable(summary(flight_delays))
```

#### 5.1.4 Checking NA count
```{r Checking NA values, include=T} 
na_count <- colSums(is.na(flight_delays))
print(na_count)
```

#### 5.1.5 Head of Cleaned data
```{r Cleaned data, include=T}
head(flight_delays)
```

#### 5.1.6 Checking all the rows if delay exists or not
```{r Delay checks, include=T}
flight_delays$outcome <- ifelse(rowSums(flight_delays == 0) == ncol(flight_delays), 1, 0)
no_delay_rows <- flight_delays[flight_delays$outcome == 1, ]
head(no_delay_rows)
```

### 5.2 Exploring the data (Question Specific)

#### 5.2.1 Frequency of top 10 Origin and Destination airports delay
```{r Identifying frequency, include=T}
origin_counts <- flight_delays %>%
  group_by(Origin) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  mutate(Type = "Origin")

dest_counts <- flight_delays %>%
  group_by(Dest) %>%
  summarise(Frequency = n()) %>%
  arrange(desc(Frequency)) %>%
  mutate(Type = "Destination")

origin_counts <- rename(origin_counts, Airport = Origin)
dest_counts <- rename(dest_counts, Airport = Dest)

combined_counts <- bind_rows(origin_counts, dest_counts)

top_airports <- combined_counts %>%
  group_by(Airport) %>%
  summarise(TotalFrequency = sum(Frequency)) %>%
  arrange(desc(TotalFrequency)) %>%
  slice(1:10)

final_counts <- combined_counts %>%
  filter(Airport %in% top_airports$Airport)

ggplot(final_counts, aes(x = reorder(Airport, Frequency), y = Frequency, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 10 Airports by Delay Frequency (Origin vs Destination)",
       x = "Airport", y = "Delay Frequency") +
  theme_minimal() +
  coord_flip() +
  scale_fill_manual(values = c("Origin" = "steelblue", "Destination" = "darkorange"))
```

Findings Summary from above Graph:

Top Delayed Airports:
ORD (Chicago O'Hare International Airport) shows the highest delay frequency for both origin and destination flights. Delays originating from ORD are slightly more frequent than those arriving at ORD.
DFW (Dallas/Fort Worth International Airport) ranks second, with delays at the destination level slightly higher than at the origin.
ATL (Hartsfield-Jackson Atlanta International Airport) ranks third, with a significant number of origin delays compared to destination delays.

#### 5.2.2 Creating Heatmap to understand delay cause at top 10 airports
```{r Identifying delay causes, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

heatmap_data <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay) %>%
  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType",
               values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  filter(!is.na(TotalDelay))

heatmap_summary <- heatmap_data %>%
  group_by(Airport, DelayType) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE)) %>%
  ungroup()

ggplot(heatmap_summary, aes(x = Airport, y = DelayType, fill = TotalDelay)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "white", high = "blue", na.value = "grey50") +
  labs(title = "Heatmap of Delay Causes at Top 10 Airports",
       x = "Airport",
       y = "Delay Cause",
       fill = "Total Delay Minutes") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Findings Summary from above Graph:

LateAircraftDelay and CarrierDelay are the two most dominant causes of delays, contributing the majority of delay minutes.
WeatherDelay and SecurityDelay have minimal impacts compared to other causes.
Efforts to reduce LateAircraftDelay and CarrierDelay should be prioritized to address the largest sources of delay.

#### 5.2.3 Identifying season in which delays occurs the most at top 10 airports
```{r Identifying seasons, include=T}
top_origin_airports <- flight_delays %>%
  group_by(Origin) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Origin)

top_dest_airports <- flight_delays %>%
  group_by(Dest) %>%
  summarize(TotalDelay = sum(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, na.rm = TRUE)) %>%
  arrange(desc(TotalDelay)) %>%
  slice_head(n = 10) %>%
  pull(Dest)

top_airports <- unique(c(top_origin_airports, top_dest_airports))

filtered_delays <- flight_delays %>%
  filter(Origin %in% top_airports | Dest %in% top_airports)

print(colnames(filtered_delays))

seasonal_delays <- filtered_delays %>%
  select(Origin, Dest, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, Season) %>%

  pivot_longer(cols = c(CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay),
               names_to = "DelayType", values_to = "TotalDelay") %>%
  mutate(Airport = ifelse(Origin %in% top_airports, Origin, Dest)) %>%
  group_by(Airport, Season) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE), .groups = 'drop')

ggplot(seasonal_delays, aes(x = Airport, y = TotalDelay, fill = Season)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Seasonal Delays for Top 10 Airports",
       x = "Airport", y = "Total Delay (Minutes)") +
  theme_minimal() +
  coord_flip() +
  scale_fill_brewer(palette = "Set3")
```

Findings Summary from above Graph:

The seasonal bar chart revealed that winter had the highest delay minutes, followed by spring. Winter delays are expected due to adverse weather conditions that impact flight operations. This seasonal analysis underscores the importance of season-specific delay mitigation strategies, such as deploying additional resources or preparing contingency plans during peak delay seasons.

### 5.3 Statistical Tests

#### 5.3.1 Two-Sample T-Test: Comparing Delays Between Two Airports
```{r T-test, include=T}
airport_data <- flight_delays %>% filter(Origin %in% c("ORD", "DFW"))

t_test_result <- t.test(
  LateAircraftDelay ~ Origin,
  data = airport_data,
  alternative = "two.sided")

print(t_test_result)
```

Implications:

This finding suggests that ORD experiences longer delays for Late Aircraft compared to DFW. Factors such as higher traffic volume, weather conditions, or airport congestion at ORD might contribute to this difference.

#### 5.3.2 ANOVA: Comparing Mean Delays Across Seasons
```{r Anova-test, include=T}
anova_result <- aov(CarrierDelay ~ Season, data = flight_delays)
summary(anova_result)
```

Conclusion:

The p-value < 0.05 indicates strong statistical evidence to reject the null hypothesis.
This result confirms that there are significant differences in the mean Carrier Delays among the seasons.

Implications:

The significant differences suggest that the season plays an important role in influencing Carrier Delays. Further analysis (e.g., post-hoc tests) can help identify which specific seasons contribute to these differences.

#### 5.3.3 Tukey HSD test for identifying which pairs of seasons differ significantly
```{r Tukey-test, include=T}
tukey_result <- TukeyHSD(anova_result)

print(tukey_result)
```

Implications:

The significant pairwise comparisons highlight that Summer consistently has lower Carrier Delays compared to both Spring and Winter, while no significant difference exists between Winter and Spring.

Now further lets make a Linear Model:

### 5.4 Linear Regression Model to Analyze Total Flight Delays Across Airports and Seasons
```{r LM, include=T}
flight_data_cleaned <- seasonal_delays %>%
  group_by(Airport, Season) %>%
  summarize(TotalDelay = sum(TotalDelay, na.rm = TRUE), .groups = 'drop')

flight_data_cleaned$Season <- factor(flight_data_cleaned$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

model <- lm(TotalDelay ~ Airport + Season, data = flight_data_cleaned)

xkabledply(model, title = paste("Model :", format(formula(model)) ) )
summary(model)
```

Key Findings:

ORD (O'Hare) shows a significant positive impact on delays, with the highest coefficient estimate (962,795).
Season Summer significantly reduces total delays (p < 0.001) compared to the reference level (Winter).
Spring does not show a statistically significant impact on total delays.

Conclusion:

The linear regression model demonstrates that delays vary significantly by airport and season, with some airports contributing more to delays and Summer generally associated with fewer delays compared to Winter. The model explains 89.1% of the variability in total delays.

```{r}
predictions <- predict(model, newdata = flight_data_cleaned)

predictions
```

### 5.5 Visualizing Predictions versus Actual values
```{r Predictions, include=T}
plot(flight_data_cleaned$TotalDelay, type = "p", col = "blue", pch = 16, cex = 1.2,
     xlab = "Observation Index", ylab = "TotalDelay",
     main = "Comparison of Actual and Predicted Total Delays")

points(predictions, col = "red", pch = 17, cex = 1.2)

legend("topright", legend = c("Actual", "Predicted"),
       col = c("blue", "red"), pch = c(16, 17), cex = 0.8)
```

Key Findings:

The comparison of actual versus predicted total flight delays is visualized in the graph, where:

Actual values are shown in blue circles, representing the observed total delays for each airport and season combination.
Predicted values are shown in red triangles, representing the delays predicted by the linear regression model.
From the plot, it is evident that the predicted values closely follow the pattern of the actual values.

In conclusion, the model appears to be robust and reliable for predicting flight delays.

```{r}
par(mfrow = c(2, 2))
plot(model)
```
### 5.6 Model Evaluation
```{r}
mse <- mean(model$residuals^2)
print(mse)

rmse <- sqrt(mse)
print(rmse)
```

Implications: 

These values indicate that the model has a reasonable level of accuracy in predicting flight delays. Although there is some error, the predictions align well with the actual values, and the model's performance is acceptable based on the calculated metrics.

Final Conclusion:

1. Key Factors Influencing Flight Delays:
Late Aircraft Delays, NAS Delays, and Carrier Delays are the primary contributors to total flight delays. Weather-related delays occur less frequently, while Security Delays are rare.
Airports such as O'Hare International (ORD), Dallas/Fort Worth International (DFW), and Hartsfield-Jackson Atlanta International (ATL) consistently experience the highest cumulative delays due to these factors.

2. Seasonal Variations:
The Winter season experiences higher delays compared to other seasons, with Summer seeing fewer delays, as confirmed by the ANOVA test and the Tukey HSD test.

3. Airport-Specific Findings:
O'Hare International (ORD) and Dallas/Fort Worth International (DFW) are outliers with consistently higher delays, as shown by the T-test results.
On the other hand, airports like JFK and LGA experience lower delays, making them more efficient in terms of total delays compared to others.

4. Linear Regression Model Insights:
Airports like JFK and LGA experience significantly lower delays compared to the baseline, while ORD experiences higher delays.
Season is a significant predictor of delays, with Summer having fewer delays than Winter. The model's Multiple R-squared value of 0.891 indicates a strong fit, suggesting that these factors together explain a significant portion of the variation in total delays.

5. Model Performance:
The Mean Squared Error (MSE) of 3.58e+10 and Root Mean Squared Error (RMSE) of 189,221 indicate a reasonable fit, with the model performing adequately in predicting total flight delays across airports and seasons.