---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala,  Hussain Nathani, Rasika Nilatkar, Sayyam Palrecha"
#date: "today"
date: "`r Sys.Date()`"
subtitle: "Team 2 - JamesRiverhouse"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r Libraries, include=T}
library(dplyr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(corrplot)
library(scales)
library(gridExtra)
library(stats)
library(reshape2)
```

# Introduction:

According to a market report by AirHelp, in 2022, at least 200 million passengers were affected by delays and cancellations in the US, which impacted on the economy of $30–34 billion in the US. With the increasing volume of air traffic, especially during peak travel seasons, understanding the causes and patterns of delays is crucial for airlines and airports to optimize resources and improve performance. This project aims to predict and analyze U.S. flight delays using data-driven techniques in R, focusing on different causes of delays, seasonal variations, and time-of-day effects. By answering key questions, we can provide insights that help airlines and airports mitigate delays and enhance operational efficiency.

Previous research on flight delays has focused on identifying key causes and attempting to forecast delays based on historical data. Studies often show weather as a major contributor, particularly in seasonal peaks, and highlight airports with chronic delays. Many existing analyses have also sought to classify delays by type, allowing airlines to target specific operational changes.

Exploring the dataset collected from Kaggle allows us to see delay patterns directly from the data. By analyzing patterns in delay causes and seasonality, we can make actionable recommendations. Our findings provide airlines and airport management with insights that could lead to better resource allocation and proactive delay management, ultimately improving the traveler experience and operational efficiency.



# 1. Loading the Dataset
```{r}
flight <- data.frame(read.csv("Flight_delay.csv"))
str(flight)
```

Dataset loaded..

```{r}
column_names <- c('Day_of_Week','Date','Departure_Time','Arrival_Time','Scheduled_Arrival_Time',  'Unique_Carrier_Code','Carrier_name', 'Flight_Number','Tail_Number', 'Actual_Elapsed_Time_min','Estimated_Elapsed_Time_min','Air_Time_min', 'Arrival_Delay_min', 'Departure_Delay_min','Origin', 'Origin_Airport','Destination','Destination_Airport', 'Distance_mi','Taxi_In_min','Taxi_Out_min', 'Cancelled', 'CancellationCode','Diverted','Carrier_Delay_min','Weather_Delay_min', 'NAS_Delay_min','Security_Delay_min','Late_Aircraft_Delay_min')

colnames(flight) <- column_names
unique_carriers <- data.frame(unique(flight[,7]))
```


# 2. Data Cleaning & Summary of Dataset

The dataset from the Kaggle link titled "Flight Delay and Causes" contains
comprehensive information on U.S. flight delays, focusing on various causes of delays
and their impacts across different airlines and airports. The data is sourced from the
U.S. Department of Transportation's Bureau of Transportation Statistics (BTS) and
includes detailed records of flight delays, cancellations, and their contributing factors.

Summary:
```{r}
summary(flight)
```

Number of rows before Data pre-processing: `r nrow(flight)`

The airlines considered for analysis are:

`r knitr::kable(unique_carriers)`


Reasons for the delay:

1. CarrierDelay → Flight delay due to carrier(e.g. maintenance or crew problems, aircraft cleaning, fueling, etc), 0 = No, yes = (in minutes)

2. WeatherDelay → Flight delay due to weather, 0 = No, yes = (in minutes)

3. NASDelay → Flight delay by NSA(National Aviation System), 0 = No, yes = (in minutes)

4. SecurityDelay → Flight delay by security reason, 0 = No, yes = (in minutes)

5. LateAircraftDelay → Flight delay by Aircraft reason, 0 = No, yes = (in minutes)


# 3. Data Preprocessing

## 3.1. Outliers and null value removal

Outlier removal improves data analysis accuracy by eliminating extreme values that can skew statistical measures and models. This leads to more reliable and interpretable results.

```{r}
remove_outliers <- function(df, columns) {
    df_clean <- df
    
    for(column_name in columns) {
        # Get the specified column
        x <- df_clean[[column_name]]
        
        # Removing NA values
        valid_rows <- !is.na(x)
        x_clean <- x[valid_rows]
        
        # Calculating Q1, Q3, and IQR
        Q1 <- quantile(x_clean, 0.25)
        Q3 <- quantile(x_clean, 0.75)
        IQR <- Q3 - Q1
        
        # bounds for outlier removal
        lower_bound <- Q1 - 1.5 * IQR
        upper_bound <- Q3 + 1.5 * IQR
        
        # summary before removal
        cat("\nProcessing", column_name, ":\n")
        cat("Original rows:", sum(valid_rows), "\n")
        cat("Q1:", Q1, "Q3:", Q3, "IQR:", IQR, "\n")
        cat("Lower bound:", lower_bound, "Upper bound:", upper_bound, "\n")
        
        # Deleting outlier rows
        keep_rows <- x >= lower_bound & x <= upper_bound & valid_rows
        df_clean <- df_clean[keep_rows, ]
        
        # summary after removal
        cat("Rows after removal:", nrow(df_clean), "\n")
        cat("Outliers removed:", sum(valid_rows) - nrow(df_clean), "\n\n")
    }
    
    return(df_clean)
}
```



```{r}
library(ggplot2)
flight_b <- remove_outliers(flight,c('Arrival_Delay_min','Carrier_Delay_min','Weather_Delay_min','NAS_Delay_min','Security_Delay_min','Late_Aircraft_Delay_min'))
```

Number of rows after nulls and Outlier removal: `r nrow(flight_b)`

## 3.2. Normality checks

Parametrical statistical testing can be done if the data points are normal. Let's check.


```{r}
#| fig-width: 20
#| fig-height: 15
par(mfrow = c(3, 2))
qqnorm(flight_b$Carrier_Delay_min, ylab = 'Flight delay due to carrier', main = "Flight delay due to carrier"); qqline(flight_b$Carrier_Delay_min,col=2)
qqnorm(flight_b$Weather_Delay_min, ylab = 'Flight delay due to weather', main = "Flight delay due to weather"); qqline(flight_b$Weather_Delay_min,col=2)
qqnorm(flight_b$NAS_Delay_min, ylab = 'NAS_Delay_min', main = "Flight delay by NSA(National Aviation System)"); qqline(flight_b$NAS_Delay_min,col=2)
qqnorm(flight_b$Security_Delay_min, ylab = 'Security_Delay_min', main = "Flight delay by security reason"); qqline(flight_b$Security_Delay_min,col=2)
qqnorm(flight_b$Late_Aircraft_Delay_min, ylab = 'Flight delay by Aircraft reason', main = "Flight delay by Aircraft reason"); qqline(flight_b$Late_Aircraft_Delay_min,col=2)
qqnorm(flight_b$Arrival_Delay_min, ylab = 'Total Delay' , main = "Total Delay"); qqline(flight_b$Arrival_Delay_min,col=2)

```


weather and security delays are rare in this dataset and when they occur, they're likely significant outliers.

None of the delay types follow a normal distribution.

Since these distributions are not normal, Parametric statistical studies cannot be performed. 


# 4. SMART Questions & EDA 

## SMART Q1
"Can we predict the total arrival delay using ML modeling for the top 5 most delayed airlines, using key factors that holds impact on the delay. This helps identify the significant predictors of delays and provides actionable insights for resource allocation and operational planning in future from the data collected from January 2019 to June 2019, catering better customer satisfaction."

### 4.1. Identifying top 5 airlines:

To Identify top 5 airlines, comparing median and mean values on boxplots would help.


```{r}
#| fig-width: 20
#| fig-height: 5
ggplot(flight_b,aes(Carrier_name,Arrival_Delay_min))+geom_boxplot(aes(fill = Carrier_name ))+labs(title = "Box plots of Airlines flight delay distribution") + theme(plot.title = element_text(size = 25,hjust =0.5)) + stat_summary(aes(y = Arrival_Delay_min , label = round(..y.., 2)), 
                 fun.y = median, geom = "text") 
```



From the box-plots above, higher median lines are found for:

JetBlue Airways

United Air Lines Inc.

American Eagle Airlines Inc.

American Airlines Inc.

US Airways Inc.


```{r}
flight_5 <- flight_b[flight_b$Carrier_name %in% c('JetBlue Airways','United Air Lines Inc.','US Airways Inc.','American Airlines Inc.','American Eagle Airlines Inc.'),]
```

```{r}
library(dplyr)

# Adding the number of flights on a date level for each origin
flight_b <- flight_b %>%
  group_by(Date, Origin) %>%
  mutate(Flights_At_Origin = n()) %>%
  ungroup()

# Adding the number of flights on a date level for each destination
flight_b <- flight_b %>%
  group_by(Date, Destination) %>%
  mutate(Flights_At_Destination = n()) %>%
  ungroup()

```

### 4.2. Identifying major reasons for these 5 airlines:

If we can understand the major reasons impacting top 5 delayed airlines, that would cover majority of the delayed reasons in the US according to the available dataset.


```{r}
library(ggplot2)
library(reshape2)
avg_delays <- aggregate(
  cbind(
    Carrier_Delay_min, 
    Weather_Delay_min, 
    NAS_Delay_min, 
    Security_Delay_min, 
    Late_Aircraft_Delay_min
  ) ~ Carrier_name, 
  data = flight_5, 
  FUN = mean, 
  na.rm = TRUE
)

delay_long <- melt(
  avg_delays,
  id.vars = "Carrier_name"
)

delay_long <- melt(avg_delays, id.vars = "Carrier_name")
delay_totals <- aggregate(value ~ Carrier_name, data = delay_long, sum)
delay_long <- merge(delay_long, delay_totals, by = "Carrier_name", 
                   suffixes = c("", "_total"))
delay_long$percentage <- (delay_long$value / delay_long$value_total)*100

threshold <- 0
delay_long$label <- ifelse(delay_long$percentage >  threshold, sprintf("%.1f%%",delay_long$percentage),"") 

```


```{r}
ggplot(delay_long, aes(x = Carrier_name, y = value, fill = variable)) +
  geom_bar(stat = "identity", position = "fill") +
  geom_text(aes(label = label),
            position = position_fill(vjust = 0.5),size = 3)+
  theme_minimal() +
  labs(
    title = "Proportion of Delay Types by Airline",
    x = "Airline",
    y = "Percentage",
    fill = "Delay Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +  
  scale_y_continuous(labels = scales::percent)

flight_5$MonthYear <- format(as.Date(flight_5$Date, "%d-%m-%Y"), "%b-%Y")

# Calculate monthly averages for each airline and delay type
monthly_delays <- aggregate(
  cbind(
    Carrier_Delay_min, 
    Weather_Delay_min, 
    NAS_Delay_min, 
    Security_Delay_min, 
    Late_Aircraft_Delay_min
  ) ~ MonthYear + Carrier_name, 
  data = flight_5, 
  FUN = mean, 
 na.rm = TRUE
)
```



```{r}
delay_long <- melt(monthly_delays, 
                  id.vars = c("MonthYear", "Carrier_name"),
                  variable.name = "Delay_Type",
                  value.name="Minutes")
```


Analysis from above graph:

1. Late Aircraft Delay dominates across all airlines, accounting for the largest proportion of delays, highest being in JetBlue Airways (64.0%).

2. Followed by Carrier Delay and NAS Delay, all three combined causes 100% delays.

3. Weather and Security delays appear minimal or non-existent

4. Late aircraft delays being dominant suggests a huge impact in airline operations

5. Different airlines might need different strategies based on their delay profiles, for example JetBlue might focus on late aircraft issues, while American Airlines might prioritize carrier-related delays since American Airlines has the highest percentage compared to others.


### 4.3. Identifying major reasons for these 5 airlines

To validate if the top5 reasons are major reasons impacting most of the flight delays, let's check the consistency of the reasons percentage across the available timeline.


```{r}
library(dplyr)

delay_long <- delay_long %>%
  mutate(
    MonthDate = as.Date(paste0("01-", MonthYear), format = "%d-%b-%Y"),
    MonthYear = factor(format(MonthDate, "%b-%Y"),levels = format(sort(unique(MonthDate)),"%b-%Y")))
ggplot(delay_long, 
       aes(x = MonthYear, 
           y = Minutes, 
           color = Delay_Type, 
           group = Delay_Type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~Carrier_name, ncol = 2, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Monthly Delay Trends by Airline and Delay Type",
    x = "Month-Year",
    y = "Average Delay Minutes",
    color = "Delay Type"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 10, face = "bold"),
    plot.title = element_text(size = 14, hjust = 0.5),
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    panel.spacing = unit(1, "lines")
  ) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(nrow=1))

```

Consistency of analysis over 6 month period:

1. Late Aircraft Delays (orange line) consistently show the highest delays across all airlines

2. Security Delays (purple line) consistently show the lowest delays

3. Most trends remain relatively stable across the 6-month period

### 4.4. Modelling:


Data preparation for modelling.

let's check correlation:

```{r}
ds <- subset(flight_5,select = -c(Carrier_Delay_min,Origin_Airport,Destination_Airport,CancellationCode,MonthYear,Carrier_name,Weather_Delay_min,NAS_Delay_min,Security_Delay_min,Late_Aircraft_Delay_min))

ds$Cancelled <- as.factor(ds$Cancelled)
ds$Diverted <- as.factor(ds$Diverted)
ds$Flight_Number <- as.factor(ds$Flight_Number)
ds$Tail_Number <- as.factor(ds$Tail_Number)
```

There are no cancelled and diverted flights, hence let's drop the columns

```{r}
ds <- subset(ds,select = -c(Cancelled,Diverted))
```


Let's convert columns into factor variables

```{r}
ds$Day_of_Week <- as.factor(ds$Day_of_Week)
ds$Date <- as.Date(ds$Date, format = "%d-%m-%Y")

ds$Unique_Carrier_Code <- as.factor(ds$Unique_Carrier_Code)
ds$Origin <- as.factor(ds$Origin)
ds$Destination <- as.factor(ds$Destination)
```

`Arrival_Time`, `Scheduled_Arrilval_Time`, `Departure Time` are left in integer format, since the time of the day are ordinal variables and has clear order.


```{r}
ds <- subset(ds,select = -c(Flight_Number,Tail_Number))
```


```{r}
loadPkg("leaps")
reg.forward10 <- regsubsets(Arrival_Delay_min~., data = ds, nvmax = 10, nbest = 2, method = "forward")
plot(reg.forward10, scale = "adjr2", main = "Adjusted R^2")
```
```{r}
reg_summary <- summary(reg.forward10)
best_model_index <- which.max(reg_summary$adjr2)  # Index of highest Adjusted R²
best_model_features <- reg_summary$which[best_model_index, ]  # Logical vector of selected features
selected_features <- names(best_model_features[best_model_features == TRUE])[-1]
cat("Features with the best Adjusted R²:\n")
print(selected_features)
ds_filtered <- ds[, c("Arrival_Delay_min", selected_features)]
```
Since the features with best Adjusted $R^2$ are `Estimated_Elapsed_Time_min`, `Air_Time_min`, `Departure_Delay_min` , `Taxi_In_min` and `Taxi_Out_min`. Let's re-check Adjusted $R^2$ graph with just these metrics.

```{r}
loadPkg("leaps")
reg.forward10 <- regsubsets(Arrival_Delay_min~., data = ds_filtered, nvmax = 10, nbest = 2, method = "forward")
plot(reg.forward10, scale = "adjr2", main = "Adjusted R^2")
```
```{r}
model <- lm(Arrival_Delay_min ~ Estimated_Elapsed_Time_min+Air_Time_min+Departure_Delay_min+Taxi_In_min+Taxi_Out_min, data = ds)
summary(model)
```

$R^2$ value of 1 may seem ideal, it’s rare in real-world datasets. This perfect fit could indicate issues such as:
Overfitting: The model may be too complex and tightly fit to the specific dataset.
Multicollinearity: There may be high correlations among the predictors, leading to redundancy.

Let's check multi collinearity for the model.
```{r}
xkablevif(model)
```
As expected, high multi - collinearity.

Hence, Let's build multiple models using the above variables.

```{r, fig.height=6, fig.width=15}
correlation_matrix <- cor(select_if(ds_filtered,is.numeric))
par(mar = c(0, 0, 0, 0), cex.main = 0.8, cex.axis = 0.8, cex.lab = 3.0)

corrplot(correlation_matrix, 
         method = "color", 
         addCoef.col = "black", 
         col = colorRampPalette(c("maroon", "yellow", "white", "lightblue", "palegreen"))(100), 
         tl.col = "black", 
         tl.cex = 0.8,          # Scale for text labels
         tl.srt = 30,           # Rotate labels for better spacing
         number.cex = 0.8,      # Scale for correlation coefficients
         mar = c(0, 0, 1, 0),   # Adjust margins (bottom, left, top, right)
         main = "Correlation plot for top 5 most delayed flight data")

```

`Departure_Delay_min` has high correlation with `Arrival_Delay_min`, which also makes sense because delay in departure also leads to flight reaching the destination late.

```{r}
model1 <- lm(Arrival_Delay_min~Departure_Delay_min,data = ds_filtered)
summary(model1)
```
$R^2$ being 74.5% and the p-value suggests that `Departure_Delay_min` is a significant factor. The F-statistic is extremely large further confirming the model's overall significance.

The second highest correlated factor is `Taxi_Out_min`.

```{r}
model2 <- lm(Arrival_Delay_min~(Departure_Delay_min*Taxi_Out_min)^2,data = ds_filtered)
summary(model2)
```
$R^2$ being 80.3% with interaction terms and the p-value suggests that `Departure_Delay_min` and `Taxi_Out_min` are a significant factor. The F-statistic is extremely large further confirming the model's overall significance.

Let's check the VIF of model2

```{r}
xkablevif(model2)
```

Vif = 1.03, Hence there's no multicollinearity.


This suggests that the time elapsed between departure from the origin airport gate and wheels off is also a significant factor to the predict the flight delay.
Extended Taxi-Out time often occurs due to congestion at the runway or delays in the sequencing of departures.
Adverse weather (rainy/winters), Insufficient ground resources (as explored in the EDA).
Taxi-Out delays peaks during holiday seasons/ winter months, de-icing procedures can extend the Taxi-Out Time, contributing to departure and arrival delays.


```{r}
model3 <- lm(Arrival_Delay_min~(Departure_Delay_min+Taxi_Out_min+Taxi_In_min)^2,data = ds_filtered)
summary(model3)
```
There's a significant improvement in $R^2$ value (82.3%) and p-values suggest that the predictors are significant.
Same as Taxi_Out delays, Taxi_In delays also peaks during holiday seasons and winter/rainy seasons.

Let's check vif values.

```{r}
xkablevif(model3)
```

VIFs until 10 can be accepted.


```{r}
model4 <- lm(Arrival_Delay_min ~ (Departure_Delay_min + Taxi_Out_min + 
    Taxi_In_min+Air_Time_min)^2, data = ds_filtered)
summary(model4)
```
Not a significant impact on the $R^2$ values.
Let's check VIF values.

```{r}
xkablevif(model4)
```

Multi collinearity kicks in when`Air_Time_min` is added.

```{r}
model5 <- lm(Arrival_Delay_min ~ (Departure_Delay_min + Taxi_Out_min + 
    Taxi_In_min+Estimated_Elapsed_Time_min)^2, data = ds_filtered)
summary(model5)
```
$R^2$ didn't improve significantly, but let's check vif
```{r}
xkablevif(model5)
```

VIF crossing 10. Hence, multi collinearity is present.

Now, let's compare model1, model2, model3 using ANOVA tests.

```{r}
anova_model <- anova(model1,model2,model3)
anova_model
```
Model Comparisons
Model 1 vs. Model 2:

Adding (Departure_Delay_min * Taxi_Out_min)^2 explains an additional 4,152,475 units of variance (Sum of Sq).
The F-statistic is 24,285, with a p-value < 2e-16, indicating this improvement is highly significant.
Model 2 vs. Model 3:

Adding Taxi_In_min and its interactions explains an additional 1,451,645 units of variance (Sum of Sq).
The F-statistic is 5,660, with a p-value < 2e-16, indicating this improvement is also highly significant.

Each step in model complexity significantly improves the fit, as shown by the large F-statistics and extremely small p-values (< 0.001).]The additional variance explained decreases from Model 2 to Model 3 (4,152,475 vs. 1,451,645). This suggests that while adding more terms improves the model, the magnitude of improvement is smaller.


Let's check MSE and RMSE for the model built using the features from the best model above:

```{r}
library(caTools)
set.seed(123) #for reproducability
split <- sample.split(ds_filtered$Arrival_Delay_min, SplitRatio = 0.8)
train_data <- subset(ds_filtered, split == TRUE)
test_data <- subset(ds_filtered, split == FALSE)
model2_train <- lm(Arrival_Delay_min ~ (Departure_Delay_min*Taxi_Out_min)^2, data = train_data)
model3_train <- lm(Arrival_Delay_min~(Departure_Delay_min+Taxi_Out_min+Taxi_In_min)^2, data = train_data)

predictions_2_test <- predict(model2_train, newdata = test_data)
predictions_3_test <- predict(model3_train, newdata = test_data)

predictions_2_train <- predict(model2_train, newdata = train_data)
predictions_2_train <- predict(model2_train, newdata = train_data)
```

```{r}
# Function to calculate MSE
calculate_mse <- function(actuals, predictions) {
  mean((actuals - predictions)^2)
}

# Function to calculate RMSE
calculate_rmse <- function(actuals, predictions) {
  sqrt(calculate_mse(actuals, predictions))
}

# Calculate MSE and RMSE for model2 (Train and Test)
mse_model2_train <- calculate_mse(train_data$Arrival_Delay_min, predictions_2_train)
rmse_model2_train <- calculate_rmse(train_data$Arrival_Delay_min, predictions_2_train)

mse_model2_test <- calculate_mse(test_data$Arrival_Delay_min, predictions_2_test)
rmse_model2_test <- calculate_rmse(test_data$Arrival_Delay_min, predictions_2_test)

# Calculate MSE and RMSE for model3 (Train and Test)
predictions_3_train <- predict(model3_train, newdata = train_data)  # Predictions for training data

mse_model3_train <- calculate_mse(train_data$Arrival_Delay_min, predictions_3_train)
rmse_model3_train <- calculate_rmse(train_data$Arrival_Delay_min, predictions_3_train)

mse_model3_test <- calculate_mse(test_data$Arrival_Delay_min, predictions_3_test)
rmse_model3_test <- calculate_rmse(test_data$Arrival_Delay_min, predictions_3_test)

# Print results
cat("Model 2 - Training MSE:", mse_model2_train, "RMSE:", rmse_model2_train, "\n")
cat("Model 2 - Testing MSE:", mse_model2_test, "RMSE:", rmse_model2_test, "\n\n")

cat("Model 3 - Training MSE:", mse_model3_train, "RMSE:", rmse_model3_train, "\n")
cat("Model 3 - Testing MSE:", mse_model3_test, "RMSE:", rmse_model3_test, "\n")

```
Both Model 2 and Model 3 demonstrate low MSE and RMSE values for both training and testing datasets, indicating they are well-fitted to the data and capable of making accurate predictions.
The RMSE values are close to each other for the training and testing sets, showing no signs of overfitting or underfitting.

The lower MSE and RMSE for Model 3 suggest it should be preferred for predicting Arrival_Delay_min.