---
title: "Predicting and Analyzing U.S. Flight Delays: Identifying Key Factors and Improving Forecast Accuracy"
Group: "JamesRiverhouse"
Members: "Hema Puchakayala, Rasika Nilatkar, Sayyam Palrecha, Hussain Nathani"
#date: "today"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

# SMART Question
**Can we predict the probability and duration of flight delays based on the time of day and type of delay, including factors like NAS, Security, and Late Aircraft?**

```{r init, include=FALSE}
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
library(ezids)
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(echo = T, warning = F, results = "markup", message = F)
options(scientific=T, digits = 3)  
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```


### Necessary Libraries
```{r Libraries, include=T}
library(dplyr)
library(knitr)
library(ggplot2)
```

### Understanding the dataset
```{r Initial dataset, include=T}
flight <- read.csv("Flight_delay.csv")
str(flight)
```

### Formatting Date & Day Column
```{r Formatting data, include=T}
#convert date column
flight$Date <- as.Date(flight$Date, format = "%d-%m-%Y")

#creating time of day column
flight <- flight %>%
  mutate(TimeOfDay = case_when(
    DepTime >= 600 & DepTime < 1200 ~ "Morning",
    DepTime >= 1200 & DepTime < 1800 ~ "Afternoon",
    DepTime >= 1800 & DepTime < 2400 ~ "Evening",
    TRUE ~ "Night"
  ))
```

### Filter for valid delay records and remove cancellations/diversions
```{r Filtering data, include=T}
filtered_data <- flight %>%
  filter(Cancelled == 0 & Diverted == 0) %>%
  select(DepDelay, CarrierDelay, WeatherDelay, NASDelay, SecurityDelay, LateAircraftDelay, TimeOfDay)
```


### Create a Binary Delay Indicator (Delayed column)
```{r Dealayed column, include=T}
filtered_data <- filtered_data %>%
  mutate(Delayed = ifelse(DepDelay > 15, 1, 0))
```


### Summary of cleaned data
```{r Summary, include=T}
knitr::kable(summary(filtered_data))
```


### Distribution of Delays by Time of Day
```{r Delays by Time of Day, include=T}
ggplot(filtered_data, aes(x = TimeOfDay, fill = factor(Delayed))) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Delays by Time of Day", x = "Time of Day", y = "Proportion", fill = "Delayed")
```

### Average Delay Duration by Delay type
```{r Delay Duration by Type, include=T}
library(tidyverse)
delay_summary <- filtered_data %>%
  summarise(
    Carrier = mean(CarrierDelay, na.rm = TRUE),
    Weather = mean(WeatherDelay, na.rm = TRUE),
    NAS = mean(NASDelay, na.rm = TRUE),
    Security = mean(SecurityDelay, na.rm = TRUE),
    LateAircraft = mean(LateAircraftDelay, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(), names_to = "DelayType", values_to = "AvgDelay")

ggplot(delay_summary, aes(x = DelayType, y = AvgDelay, fill = DelayType)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Delay Duration by Type", x = "Delay Type", y = "Average Delay (Minutes)")
```

### Sample t-test
```{r t-test, include=T}
# Subset data for Morning and Evening
morning_delays <- filtered_data %>% filter(TimeOfDay == "Morning") %>% pull(DepDelay)
evening_delays <- filtered_data %>% filter(TimeOfDay == "Evening") %>% pull(DepDelay)

# Perform a t-test
t_test_result <- t.test(morning_delays, evening_delays, alternative = "two.sided")
print(t_test_result)
```
### ANOVA test
```{r ANOVA, include=T}
 # Perform ANOVA
anova_result <- aov(DepDelay ~ TimeOfDay, data = filtered_data)
summary(anova_result)
```
### Tukey HSD test
```{r Turkey HSD test, include=T}
# Perform Tukey HSD test
tukey_result <- TukeyHSD(anova_result)
print(tukey_result)

# Visualize Tukey HSD results
plot(tukey_result)
```

# Predictive Modelling

### Logistic Regression for Delay Probability
```{r Logistic regression, include=T}
# Split into train and test datasets
set.seed(123)
train_idx <- sample(nrow(filtered_data), 0.7 * nrow(filtered_data))
train_data <- filtered_data[train_idx, ]
test_data <- filtered_data[-train_idx, ]

# Logistic regression model
logit_model <- glm(Delayed ~ TimeOfDay + NASDelay + SecurityDelay + LateAircraftDelay,
                   data = train_data, family = binomial(link = "logit"))

# Model summary
summary(logit_model)
```

### Evaluation of Logistic Regression
```{r}
### Evaluation of Logistic Regression:
library(caret)
library(pROC)

# Predictions
logit_preds <- predict(logit_model, test_data, type = "response")  # Use logit_model
logit_preds_class <- factor(if_else(logit_preds > 0.5, 1, 0), levels = c(0, 1))

# Confusion Matrix
confusion_matrix <- confusionMatrix(logit_preds_class, factor(test_data$Delayed))
print(confusion_matrix)

# ROC Curve and AUC
roc_curve <- roc(as.numeric(test_data$Delayed), logit_preds)
auc_value <- auc(roc_curve)

# Plot ROC curve
plot(roc_curve, main = "ROC Curve for Logistic Regression")
cat("AUC:", auc_value)
```

### Linear Regression: Predicting Delay Duration

```{r Linear Regression, include=T}
# Filter for delayed flights only
delayed_flights <- train_data %>% filter(Delayed == 1)

# Linear regression model
lm_model <- lm(LateAircraftDelay ~ TimeOfDay + NASDelay + SecurityDelay, data = delayed_flights)

# Model summary
summary(lm_model)
```

### Evaluation of Linear Regression
```{r include=T}
# Predictions
lm_preds <- predict(lm_model, test_data)

# Metrics
rmse <- sqrt(mean((lm_preds - test_data$LateAircraftDelay)^2, na.rm = TRUE))
r_squared <- summary(lm_model)$r.squared

cat("RMSE:", rmse, "\nR-squared:", r_squared)
```



# Final Conclusion

1. Key Predictors of Flight Delays: Logistic regression analysis identified significant predictors for flight delays, including Time of Day, NASDelay, SecurityDelay, and LateAircraftDelay. These factors provide actionable insights for airline operations to anticipate and mitigate delays.

2. Time of Day Impacts Delays: Analysis showed clear variations in delay probability and duration based on the time of day. Flights scheduled in the evening showed a higher likelihood of delays compared to other time periods.

3. Delay Types and Their Contributions: Among the delay types analyzed, Late Aircraft Delays and NAS Delays contributed the most to overall average delay duration. These categories should be prioritized for operational improvements.

Statistical Insights:

1. T-test Results: Significant differences in average departure delays were found between morning and evening flights, confirming time-based variations.
2. ANOVA and Tukey HSD: Statistical tests revealed that delay durations varied significantly across different times of the day, supporting the observed patterns.

Logistic Regression Model Performance:

1. The logistic regression model achieved a good predictive performance with a clear ROC curve and AUC score. The confusion matrix provided insights into classification accuracy.

2. Linear Regression for Delay Duration Prediction: The linear regression model effectively predicted delay durations for delayed flights. Metrics like RMSE and R-squared indicated the model’s reliability in forecasting delay times based on specific factors.

Recommendations for Airlines:

1. Focus on optimizing evening flight schedules to minimize delays.
2. Address operational inefficiencies contributing to Late Aircraft Delays and NAS Delays.
3. Use predictive models to proactively manage high-risk flights and allocate resources effectively.

This comprehensive analysis provides actionable insights for improving U.S. flight schedules, delay management, and passenger satisfaction while contributing to predictive modeling in the aviation domain.
