---
title: "Intro to DS - Midterm"
author: "Sayam Palrecha"
date: "`r Sys.Date()`"
# date: 
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r}
library(ezids)
knitr::opts_chunk$set(warning = F, results = "hide", message = F)
options(scientific=T, digits = 3) 
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(corrplot)
library(scales)
```

# SMART Question
What are the top 3 carriers with the highest average delay times during peak summer months (June-August), and how do their delay patterns correlate with specific routes and time of day

# Unserstanding the dataset
```{r}
airline_data <- read.csv('Flight_delay.csv')
str(airline_data)
```

```{r}
summary(airline_data)
```

```{r}
colSums(is.na(airline_data))
```
## Formatting the date and month column
```{r}

airline_data$Date <- as.Date(airline_data$Date)
airline_data$Month <- month(airline_data$Date)
```
# Filter summer months (June-August)
```{r}
summer_data <- airline_data %>%filter(Month %in% c(6, 7, 8))
```
## Calculate average delays by carrier
```{r}

carrier_delays <- summer_data %>%
  group_by(UniqueCarrier) %>%
  summarise(
    AvgArrDelay = mean(ArrDelay, na.rm = TRUE),
    AvgDepDelay = mean(DepDelay, na.rm = TRUE),
    TotalDelay = AvgArrDelay + AvgDepDelay,
    CarrierDelay = mean(CarrierDelay, na.rm = TRUE),
    WeatherDelay = mean(WeatherDelay, na.rm = TRUE)
  ) %>%
  arrange(desc(TotalDelay))
```
# Get top 3 carriers
```{r}
top_3_carriers <- head(carrier_delays, 3)
```
## Visualization 1: Top 3 Carriers Average Delays
```{r}

ggplot(top_3_carriers, aes(x = reorder(UniqueCarrier, -TotalDelay), y = TotalDelay)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_text(aes(label = round(TotalDelay, 2)), vjust = -0.5) +
  theme_minimal() +
  labs(
    title = "Top 3 Carriers with Highest Average Delays (Summer Months)",
    x = "Carrier",
    y = "Average Total Delay (minutes)"
  )
```
## Time of Day Analysis
```{r}

summer_data$Hour <- as.numeric(format(strptime(sprintf("%04d", summer_data$DepTime),format = "%H%M"), "%H"))
```
# Visualization 2: Delay Patterns by Time of Day
```{r}
# Visualization 2: Delay Patterns by Time of Day
hourly_delays <- summer_data %>%
  filter(UniqueCarrier %in% top_3_carriers$UniqueCarrier) %>%
  group_by(UniqueCarrier, Hour) %>%
  summarise(AvgDelay = mean(ArrDelay, na.rm = TRUE))

ggplot(hourly_delays, aes(x = Hour, y = AvgDelay, color = UniqueCarrier)) +
  geom_line() +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(
    title = "Delay Patterns by Time of Day",
    x = "Hour of Day",
    y = "Average Arrival Delay (minutes)"
  )
```
## Route Analysis
```{r}
 
route_delays <- summer_data %>%
  group_by(UniqueCarrier, Origin, Dest) %>%
  summarise(
    AvgDelay = mean(ArrDelay, na.rm = TRUE),
    Flights = n()
  ) %>%
  filter(Flights >= 50) %>%  # Filter routes with minimum 50 flights
  arrange(desc(AvgDelay))
```
# Visualization 3: Top 10 Routes with Highest Delays
```{r}

top_10_routes <- head(route_delays, 10)
ggplot(top_10_routes, 
       aes(x = paste(Origin, "->", Dest), y = AvgDelay, fill = UniqueCarrier)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Routes with Highest Average Delays",
    x = "Route",
    y = "Average Delay (minutes)"
  )
```
# Correlation Analysis
```{r}

correlation_vars <- summer_data %>%
  select(ArrDelay, DepDelay, Distance, TaxiIn, TaxiOut)
correlation_matrix <- cor(correlation_vars, use = "complete.obs")

# Visualization 4: Correlation Heatmap
corrplot(correlation_matrix, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)
```
# Statistical Summary
```{r}


summary_stats <- summer_data %>%
  group_by(UniqueCarrier) %>%
  filter(UniqueCarrier %in% top_3_carriers$UniqueCarrier) %>%
  summarise(
    Mean_Delay = mean(ArrDelay, na.rm = TRUE),
    Median_Delay = median(ArrDelay, na.rm = TRUE),
    SD_Delay = sd(ArrDelay, na.rm = TRUE),
    Total_Flights = n(),
    Delayed_Flights = sum(ArrDelay > 0, na.rm = TRUE),
    Delay_Rate = Delayed_Flights / Total_Flights * 100
  )
print(summary_stats)
```
```{r}
write.csv(summary_stats, "airline_delay_analysis.csv")
```

